### 2018-05-27 1_intro.tex
Cleanup finale up to Xevent.

\begin{figure}[t]
\centering
\ttfamily
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}[b]{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MOV EAX,[y] & MOV EBX,[x] \\
& \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: allowed} \tabularnewline
\hline
\end{tabular}
\caption{Without synchronisation}
\label{simple_wmm_x86:nosync}
\end{subfigure}
\hfill
%
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MFENCE & MFENCE \\
MOV EAX,[y] & MOV EBX,[x] \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: forbidden} \tabularnewline
\hline
\end{tabular}
\caption{With synchronisation}
\label{simple_wmm_x86:sync}
\end{subfigure}
%
\caption{Store buffering~(SB): A litmus test illustrating the write-read reordering allowed by the x86-TSO memory model}
\label{simple_wmm_x86}
\end{figure}

@@

Research on weak memory models firstly aims to \textit{formalise} an approach of understanding programs with respect to weak memory models, which is \textit{systematic}, \textit{sound} and \textit{complete}.
One of the most well-known frameworks for weak memory model-aware analysis was formalised by J. Alglave in~2010~\cite{alglave2010shared}.
It is the event-based non-deterministic model without global time (see Section~\ref{ch:wmm:event} for details).

In addition to developing the theoretical basis, researchers work on extracting the memory models for hardware architectures from existing implementations or from the specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over the last decade, memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architectures~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, etc. %, or Alpha architecture~\cite{alphaWMM}~, and others.
There are projects for validating hardware architectures wrt. a memory model as well, e.g.,~\cite{lustig2014pipecheck,lustig2016coatcheck}.

@@

The very first tool that performs the WMM-aware portability analysis is \porthos{}%
%
\footnote{The \porthos{} project web site: \url{http://github.com/hernanponcedeleon/PORTHOS}} %
%
introduced in April~2017~\cite{Porthos17a}.
This tool reduces described problem to a bounded reachability problem, which can be solved via an SMT solver.
This approach allows to capture symbolically the semantics of analysing program and both weak memory models into a single SMT-formula, augmented by the reachability assertion.
As most modern SMT solvers are efficient enough to be able to operate the state space of size millions of variables bounded by millions of constraints~(\cite{malik2009boolean}), the used method can be applicable in solving the real-world problems.
% about SAT: https://courses.cs.washington.edu/courses/csep573/11wi/lectures/ashish-satsolvers.pdf

>>>>

\begin{figure}[t]
\centering
\ttfamily
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}[b]{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MOV EAX,[y] & MOV EBX,[x] \\
& \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: allowed} \tabularnewline
\hline
\end{tabular}
\caption{Without synchronisation}
\label{simple_wmm_x86:nosync}
\end{subfigure}
\hfill
%
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MFENCE & MFENCE \\
MOV EAX,[y] & MOV EBX,[x] \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: forbidden} \tabularnewline
\hline
\end{tabular}
\caption{With synchronisation}
\label{simple_wmm_x86:sync}
\end{subfigure}
%
\caption{Store buffering~(SB): A litmus test illustrating the write-read reordering allowed by the \texttt{x86-TSO} memory model}
\label{simple_wmm_x86}
\end{figure}

@@

Research on weak memory models firstly aims to \textit{formalise} an approach of understanding programs with respect to weak memory models, which is \textit{systematic}, \textit{sound} and \textit{complete}.
One of the most well-known frameworks for weak memory model-aware analysis was formalised by J. Alglave in~2010~\cite{alglave2010shared}.
It is the event-based non-deterministic model without global time (see Section~\ref{ch:wmm:model} for details).

In addition to developing the theoretical basis, researchers work on extracting the memory models for hardware architectures from existing implementations or from the specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over the last decade, memory models have been defined for most mainstream multiprocessor architectures, such as \texttt{x86-TSO} and \texttt{Sparc-TSO} (for \textit{Total Store Order}) model for x86 and Sparc architectures~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, etc. %, or Alpha architecture~\cite{alphaWMM}~, and others.
There are projects for validating hardware architectures wrt. a memory model as well, e.g.,~\cite{lustig2014pipecheck,lustig2016coatcheck}.

@@

The very first tool that performs the WMM-aware portability analysis is \porthos{}%
%
\footnote{The \porthos{} project web site: \url{http://github.com/hernanponcedeleon/PORTHOS}} %
%
introduced in April~2017~\cite{Porthos17a}.
This tool reduces described problem to a bounded reachability problem, which can be solved via an SMT-solver (see Section~\ref{ch:enc:bmc}).
This approach allows to capture symbolically the semantics of analysing program and both weak memory models into a single SMT-formula, augmented by the reachability assertion.
As most modern SMT-solvers are efficient enough to be able to operate the state space of size millions of variables bounded by millions of constraints~(\cite{malik2009boolean}), the used method can be applicable in solving the real-world problems.
% about SAT: https://courses.cs.washington.edu/courses/csep573/11wi/lectures/ashish-satsolvers.pdf

### 2018-05-27 2_wmm.tex
Cleanup finale up to Xevent.

\section{Event-based program representation}
\label{ch:wmm:event}

@@

Thus, there are only four possible executions defined by the choice of \rf-relation.
The candidate executions pictured in Figures~\ref{simple_wmm_x86_pic:sub1}--\ref{simple_wmm_x86_pic:sub3} are consistent both under strong memory model SC and under relaxed memory models x86-TSO, Power, ARM, and some others.
However, the execution shown in Figure~\ref{simple_wmm_x86_pic:sub3} is still consistent under relaxed-memory architectures, but it becomes inconsistent under SC architecture as it forbids cycles over~$\fr\cup\po$.
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

For instance, the \fr-relation is defined as a sequence of inverted \rf-relation and \co-relation: $\texttt{fr} = (\invrel{rf};\texttt{co})$.
As an example of memory model definition in \cat{} language, Figure~\ref{ex:x86-cat} presents the excerpt from the x86-TSO memory model~\cite{herd10tutorial}.
This memory model specification asserts  acyclicity of the communication relation (the union of \rf-, \fr- and \co- relations), \texttt{po-loc} -relation, \texttt{mfence}-relation and some other derived relations~\cite{owens2009better}.

@@

empty atom & (fre;coe)
acyclic po-loc | com
acyclic GHB
\end{lstlisting}
\caption{Excerpt from the x86-TSO memory model in the \cat{} language}
\label{ex:x86-cat}
\end{figure}

>>>>

\section{Event-based program representation}
\label{ch:wmm:model}

@@

Thus, there are only four possible executions defined by the choice of \rf-relation.
The candidate executions pictured in Figures~\ref{simple_wmm_x86_pic:sub1}--\ref{simple_wmm_x86_pic:sub3} are consistent both under strong memory model SC and under relaxed memory models \texttt{x86-TSO}, \texttt{POWER}, \texttt{ARM}, and some others.
However, the execution shown in Figure~\ref{simple_wmm_x86_pic:sub3} is still consistent under relaxed-memory architectures, but it becomes inconsistent under SC architecture as it forbids cycles over~$\fr\cup\po$.
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

For instance, the \fr-relation is defined as a sequence of inverted \rf-relation and \co-relation: $\texttt{fr} = (\invrel{rf};\texttt{co})$.
As an example of memory model definition in \cat{} language, Figure~\ref{ex:x86-cat} presents the excerpt from the \texttt{x86-TSO} memory model~\cite{herd10tutorial}.
This memory model specification asserts  acyclicity of the communication relation (the union of \rf-, \fr- and \co- relations), \texttt{po-loc} -relation, \texttt{mfence}-relation and some other derived relations~\cite{owens2009better}.

@@

empty atom & (fre;coe)
acyclic po-loc | com
acyclic GHB
\end{lstlisting}
\caption{Excerpt from the \texttt{x86-TSO} memory model in the \cat{} language}
\label{ex:x86-cat}
\end{figure}
